---
title: "spaTMB"
author: "Berent Lunde"
date: "5 april 2019"
output:
  html_document:
    toc: true
    toc_depth: 3
---

\newcommand{\sp}{\hat{\mathbf{s}}(\theta)}
\newcommand{\u}{\hat{\mathbf{u}}(\theta)}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We seek to employ the framework of the R package Template Model Builder (**TMB**)-ref- ++ -- which approximately "integrates out" latent variables using the Laplace approximation -- to automatically solve the inner problem of the saddlepoint approximation (SPA) and return the negative logarithm of the SPA.
The table of contents gives an idea of how the document is laid out.


### Introduction to the Laplace approximation

We consider the Laplace approximation in the context of likelihood estimation, following (read: stolen from) -TMB paper- completely:
Let $f(\mathbf{u},\theta)$  denote the negative joint log-likelihood of the data and the random effects. This
depends on the unknown random effects $\mathbf{u}\in\mathbb{R}^n$ and parameters $\theta\in\mathbb{R}^m$.
The maximum likelihood estimate for θ maximizes
$$
\begin{align*}
    L(\theta) = \int_{\mathbb{R}^n} \exp(-f(\mathbf{u},\theta)) d\mathbf{u},
\end{align*}
$$
i.e., the random effects $\mathbf{u}$ have been integrated out.
High-dimensional integration is, in general, difficult, 
but can be achieved by applying the Laplace approximation,
giving the approximation
$$
\begin{align*}
    L^* = \frac{\sqrt{2\pi}^n \exp(-f(\mathbf{\u},\theta))}{\sqrt{\| \nabla_\mathbf{u}^2f(\mathbf{u},\theta) \|}},   
\end{align*}
$$
where

- $\u = \arg\min_\mathbf{u} f(\mathbf{u},\theta)$,
- and $\nabla_\mathbf{u}^2f(\mathbf{u},\theta)$ is the Hessian matrix, denoted $H(\theta)$ from here on out.



### Introduction to the Saddlepoint approximation

The SPA takes as a vantage point the following inversion integral:
$$
\begin{align*}
    f_\mathbf{Y}(\mathbf{y};\theta) = \frac{1}{(2\pi)^n} \int_{\mathbb{R}^n} e^{( (\tau + i \mathbf{s})^T  \mathbf{y} ) } 
    M_\mathbf{Y}(\tau + i\mathbf{s} ; \theta) d\mathbf{s}.
\end{align*}
$$
where $i=\sqrt{-1}$, $\tau\in\mathbb{R}^n$ such that $E(\exp(\tau^T\mathbf{y})) < \infty$. - Butler

Applying the Laplace approximation to this integral, we obtain multivaraite saddlepoint approximation: 
$$
\texttt{spa}(f;\mathbf{y}) = 
\frac{\exp\left(K_\mathbf{y}(\sp;\theta)-\sp^T \mathbf{y}\right)}
{(2\pi)^{(n/2)}\sqrt{\| \nabla_\mathbf{s}^2 K_\mathbf{y}(\sp;\theta) \|}}
$$

where

- $K_\mathbf{y}(\mathbf{s};\theta)  = \log M_\mathbf{y}(\mathbf{s};\theta)$,
- $\nabla_\mathbf{s}^2 K_\mathbf{y}(\mathbf{s};\theta)$ is the Hessian matrix, 
- and the saddlepoint $\sp$ solves $\sp = \arg\min_{\mathbf{s}} K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y}$

see Kleppe-Skaug 2006 for a general derivation.


### Introduction to **TMB**


Template model builder, or **TMB** for short, wraps (on the C++ side) the high-performance C++ libraries **Eigen** (a C++ template library for linear algebra: matrices, vectors, numerical solvers, and related algorithms) and **CppAD** (templated automatic differentiation using operator overloading), for R users to obtain highly efficient evaluations of objective functions and their derivatives (gradient and Hessian) and with automatically performing the Laplace approximation as its high-point.

The **TMB** user only has to specify the objective function $f(\theta)$, or $f(\mathbf{u}, \theta)$ in the case of latent variables $\mathbf{u}$ that needs to be integrated out, and **TMB** takes care of the rest.
Specifically, the (negative log) Laplace approximation to the negative joint log-likelihood is returned:
$$
 - \log L^*(\theta) = -n \log \sqrt{2\pi} + \frac{1}{2}\log \| H(\theta) \| + f(\u,\theta).
$$
Using this, the likelihood estimates of $\theta$, i.e. $\hat{\theta}=\arg\min_\theta$, can be found very efficiently, because **TMB** also returnes the gradient and Hessian of $-\log L^*(\theta)$ w.r.t. $\theta$.

#### Example from TMB

<details> <summary> R code</summary>
```{r tmb, eval=F, tidy=T}
## Simulate data
set.seed(123)
n <- 10000
sigma <- .3
phi <- .8
simdata <- function(){
    u <- numeric(n)
    u[1] = rnorm(1)
    if(n>=2)
        for(i in 2:n){
            u[i] = phi * u[i-1] + rnorm(1, sd = sqrt(1 - phi^2))
        }
    u <- u * sigma
    x <- as.numeric( rbinom(n, 1, plogis(u)) )
    data <- list(obs=x)
    data
}
##
data <- simdata()
parameters <- list(phi=phi, logSigma=log(sigma))

## Adapt parameter list
parameters$u <- rep(0,n)

require(TMB)
compile('laplace.cpp')
dyn.load(dynlib('laplace'))

obj <- MakeADFun(data, parameters, random="u", DLL="laplace")
obj$fn()

system.time( opt <- nlminb(obj$par,obj$fn,obj$gr) )
(sdr <- sdreport(obj))

```
</details>
<details> <summary> C++ code</summary>
```{r tmbcpp, eval=F, tidy=T}
#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator() ()
{
  DATA_VECTOR(obs);
  PARAMETER(phi);
  PARAMETER(logSigma);
  PARAMETER_VECTOR(u);
  using namespace density;
  Type sigma= exp(logSigma);
  Type f = 0;
  f += SCALE(AR1(phi), sigma)(u);
  vector<Type> p = invlogit(u);
  f -= dbinom(obs, Type(1), p, true).sum();
  return f;
}

```
</details>


### Using **TMB** in numerical SPA calculations and parameter optimization

Set the goal of making **TMB** automatically solve the inner problem of the SPA, 
and returning the negative log SPA for the purpose of likelihood optimisation.
Ideally, we would have **TMB** return
$$
    -\log \texttt{spa}(f,\mathbf{y})
    = \frac{1}{2}\left( \log \| \nabla_\mathbf{s}^2 K_\mathbf{y}(\sp;\theta) \| + n\log(2\pi) \right) - 
    \left(K_\mathbf{y}(\sp;\theta)-\sp^T \mathbf{y}\right).
$$
However, to solve the inner problem of the SPA, we need to supply the following objective:
$$
\begin{align*}
f(\mathbf{u},\theta) &= K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y} - n\log (2\pi)\\
&\sim K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y},
\end{align*}
$$
where $\mathbf{s}$ is set to $\texttt{random}$ (and hence, will be integrated out by **TMB**).
However, **TMB** would then return
$$
- \log L^*(\theta) = -\log \texttt{spa}(f,\mathbf{y}) + 2 \left(K_\mathbf{y}(\sp;\theta)-\sp^T \mathbf{y}\right).
$$
i.e., the sign is reversed on $f(\mathbf{u},\theta) \sim K_\mathbf{y}(\sp;\theta)-\sp^T \mathbf{y}$.
This is because, for the SPA, the Laplace approximation is calculated w.r.t. the objective
$$
f(\mathbf{u},\theta) = K_\mathbf{y}(\mathbf{\tau} + i\mathbf{s};\theta) - (\mathbf{\tau} + i\mathbf{s})^T\mathbf{y} - n\log (2\pi),
$$
where the imaginary unit, $i$, reverses the sign of the Hessian (w.r.t $\mathbf{\tau}$), replacing minimization with maximization in the the inner optimization and vice versa.


The solution is however simple:
supply to **TMB** the objective function
$$
f(\mathbf{u},\theta) = K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y} - n\log (2\pi),
$$
so that the inner problem is solved,
but supply the negative objective, gradient and (possibly) Hessian to the outer optimisation routine in R.
**TMB** will then return
$$
    -\log L^*(\theta) = 
    =
    \texttt{spa}
$$
and derivatives of the objective.

The alternative, to make everything completely automatic, would be to switch the inner optimization from minimization to maximization, and supply the negative of the proposal.

To make use of **TMB** in calculating the SPA



lines 578-- at https://github.com/kaskr/adcomp/blob/master/TMB/R/TMB.R



Then, ideally, we would supply the original inner problem of the SPA:

where $\mathbf{\tau}$ is set to $\texttt{random}$.
The effect of the imaginary unit is to reverse the sign of the Hessian (w.r.t $\mathbf{\tau}$), 

**TMB** does not directly allow complex AD data types (while still possible inside TMB, see e.g. [this file on Github](https://github.com/Blunde1/it-ift/blob/master/implementation/includes/complex.hpp) and example use (advanced SPA calculations) in corresponding repository).

The solution is however simple:
supply to **TMB** the objective function
$$
f(\mathbf{u},\theta) = K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y} - n\log (2\pi),
$$
so that the inner problem is solved,
but supply the negative objective, gradient and (possibly) Hessian to the outer optimisation routine in R.
**TMB** will then return
$$
    -\log L^*(\theta) = 
    =
    \texttt{spa}
$$
and derivatives of the objective.

The alternative, to make everything completely automatic, would be to switch the inner optimization from minimization to maximization, and supply the negative of the proposal.

To make use of **TMB** in calculating the SPA

From the introductions of the Laplace approximation, **TMB** and the SPA, it should be clear
that if we supply the inner problem of the SPA
$$
f(\mathbf{u},\theta) = K_\mathbf{y}(\mathbf{s};\theta) - \mathbf{s}^T\mathbf{y},
$$
as the objective function in **TMB**, and choose $\mathbf{s}(\theta)$ as $\texttt{random}$,
then **TMB** will return

and corresponding derivatives.
Using this, we can supply ... to some optimization algorithm and obtain the likelihood estimates $\hat{\theta}$.



### **spaTMB** example